---
title: "FARS Data Analysis "
author: "Pradeep S"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

FARS is a nationwide census providing NHTSA, Congress and the American public yearly data regarding fatal injuries suffered in motor vehicle traffic crashes.


## Working with FARS

You can leverage this package to plot maps showing the points where accidents have been reported.  

## Getting Started 

This package consists of the following methods. We will describe how these methods are used in the package.

Let's first load the packages.

```{r load-packages, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

### fars_read

This function reads a csv file and creates a data frame. This function requires packages - deplyr and readr.If the file name is not present, the function throws errors.

Let us see how this function can be used to load the data.

```{r echo=TRUE, results='markup'}
fars_read <- function(filename) {
        if(!file.exists(filename))
                stop("file '", filename, "' does not exist")
        data <- suppressMessages({
                readr::read_csv(filename, progress = FALSE)
        })
        dplyr::tbl_df(data)
}

dat <- fars_read("accident_2013.csv.bz2")
dat
```


### make_filename

Given a year, this function builds a file name string which can then be used by fars_read function to read the file.

```{r echo=TRUE, results='markup'}
make_filename <- function(year) {
        year <- as.integer(year)
        sprintf("accident_%d.csv.bz2", year)
}

filename <- make_filename(2013)
filename
```

You can see from above that calling this function with 2013 gives you the filename accident_2013.csv.bz2.

### fars_read_years

Given a year or a vector of years this function builds data frame with columns MONTH and "year"

```{r echo=TRUE, results='markup'}
fars_read_years <- function(years) {
        lapply(years, function(year) {
                file <- make_filename(year)
                tryCatch({
                        dat <- fars_read(file)
                        dplyr::mutate(dat, year = year) %>%
                                dplyr::select(MONTH, year)
                }, error = function(e) {
                        warning("invalid year: ", year)
                        return(NULL)
                })
        })
}

dat <- fars_read_years(2013)
dat
```

### fars_summarize_years

This function returns a summary data frame of number of accidents per month for a given vector of years.

```{r echo=TRUE, results='markup'}
fars_summarize_years <- function(years) {
        dat_list <- fars_read_years(years)
        dplyr::bind_rows(dat_list) %>%
                dplyr::group_by(year, MONTH) %>%
                dplyr::summarize(n = n()) %>%
                tidyr::spread(year, n)
}

dat <- fars_summarize_years(2013)
dat
```

You will relalize that the column name itself is now the year in string '2013'.

Note what happens if you call the above function by passing all the three years in a vector.

```{r echo=TRUE, results='markup'}
dat <- fars_summarize_years(c(2013,2014,2015))
dat
```

### fars_map_state

This function plots a map indicating the points where the accidents have been reported. Let us see how this function returns a map for state number 6 and the year 2013.

```{r echo=TRUE, results='markup'}
fars_map_state <- function(state.num, year) {
        filename <- make_filename(year)
        data <- fars_read(filename)
        state.num <- as.integer(state.num)

        if(!(state.num %in% unique(data$STATE)))
                stop("invalid STATE number: ", state.num)
        data.sub <- dplyr::filter(data, STATE == state.num)
        if(nrow(data.sub) == 0L) {
                message("no accidents to plot")
                return(invisible(NULL))
        }
        is.na(data.sub$LONGITUD) <- data.sub$LONGITUD > 900
        is.na(data.sub$LATITUDE) <- data.sub$LATITUDE > 90
        with(data.sub, {
                maps::map("state", ylim = range(LATITUDE, na.rm = TRUE),
                          xlim = range(LONGITUD, na.rm = TRUE))
                graphics::points(LONGITUD, LATITUDE, pch = 46)
        })
}

fars_map_state(6,2013)
```

